# Comparing Means/Quantities using Two Independent Samples {#CI-Indep-Sample-Means}

Here, we'll consider the problem of estimating a confidence interval to describe the difference in population means (or medians) based on a comparison of two samples of quantitative data, gathered using an independent samples design. Specifically, we'll look at the randomized controlled trial of Ibuprofen in Sepsis patients described in Chapter \@ref(Sepsis_RCT). 

In that trial, 300 patients meeting specific criteria (including elevated temperature) for a diagnosis of sepsis were randomly assigned to either the Ibuprofen group (150 patients) and 150 to the Placebo group. Group information (our exposure) is contained in the `treat` variable. The key outcome of interest to us was `temp_drop`, the change in body temperature (in $^{\circ}$C) from baseline to 2 hours later, so that positive numbers indicate drops in temperature (a good outcome.) Here's the comparison of `temp_drop` summary statistics in the two `treat` groups.

```{r}
mosaic::favstats(temp_drop ~ treat, data = sepsis)
```

## t-based CI for population mean difference $\mu_1 - \mu_2$ from Independent Samples

### The Pooled t procedure

The most commonly used t-procedure for building a confidence interval assumes not only that each of the two populations being compared follows a Normal distribution, but also that they have the same population variance. This is the pooled t-test, and it is what people usually mean when they describe a two-sample t test.

```{r}
sepsis %$% t.test(temp_drop ~ treat,
                  conf.level = 0.90,
                  alt = "two.sided",
                  var.equal = TRUE)
```

Or, we can use `tidy` on this object:

```{r}
tt1 <- sepsis %$% t.test(temp_drop ~ treat,
                  conf.level = 0.90,
                  alt = "two.sided",
                  var.equal = TRUE)
tidy(tt1)
```


### Using linear regression to obtain a pooled t confidence interval

A linear regression model, using the same outcome and predictor (group) as the pooled t procedure, produces the same confidence interval, again, under the assumption that the two populations we are comparing follow a Normal distribution with the same (population) variance.

```{r}
model1 <- lm(temp_drop ~ treat, data = sepsis)

tidy(model1, conf.int = TRUE, conf.level = 0.90)
```

We see that our point estimate from the linear regression model is that the difference in `temp_drop` is `r coef(model1)[2]`, where Ibuprofen subjects have higher `temp_drop` values than do Placebo subjects, and that the 90% confidence interval for this difference ranges from `r confint(model1, level = 0.90)[2]` to `r confint(model1, level = 0.90)[4]`.

We can obtain a t-based confidence interval for each of the parameter estimates in a linear model directly using `confint`. Linear models usually summarize only the estimate and standard error. Remember that a reasonable approximation in large samples to a 95% confidence interval for a regression estimate (slope or intercept) can be obtained from estimate $\pm$ 2 * standard error.

```{r}
summary(model1)
```

So, in the case of the `treatPlacebo` estimate, we can obtain an approximate 95% confidence interval with `r round(coef(summary(model1))["treatPlacebo",1],3)` $\pm$ 2 x `r round(coef(summary(model1))["treatPlacebo",2],3)` or (`r round(coef(summary(model1))["treatPlacebo",1] - 2*coef(summary(model1))["treatPlacebo",2],3)`, `r round(coef(summary(model1))["treatPlacebo",1]+2*coef(summary(model1))["treatPlacebo",2],3)`). Compare this to the 95% confidence interval available from the model directly, shown in the tidied output above, or with the `confint` command below, and you'll see only a small difference.

```{r}
confint(model1, level = 0.95)
```

### The Welch t procedure

The default confidence interval based on the t test for independent samples in R uses something called the Welch test, in which the two populations being compared are not assumed to have the same variance. Each population is assumed to follow a Normal distribution. 

```{r}
sepsis %$% t.test(temp_drop ~ treat,
                  conf.level = 0.90, 
                  alt = "two.sided")
```

Tidying works in this situation, too.

```{r}
tt0 <- sepsis %$% t.test(temp_drop ~ treat,
                  conf.level = 0.90, 
                  alt = "two.sided")

tidy(tt0)
```

## Bootstrap CI for $mu_1 - \mu_2$ from Independent Samples

The `bootdif` function contained in the `Love-boost.R` script, that we will use in this setting is a slightly edited version of the function at http://biostat.mc.vanderbilt.edu/wiki/Main/BootstrapMeansSoftware. Note that this approach uses a comma to separate the outcome variable (here, `temp_drop`) from the variable identifying the exposure groups (here, `treat`).

```{r, message = FALSE}
set.seed(431212)

sepsis %$% bootdif(temp_drop, treat, conf.level = 0.90)
```


## Wilcoxon-Mann-Whitney "Rank Sum" CI from Independent Samples

As in the one-sample case, a rank-based alternative attributed to Wilcoxon (and sometimes to Mann and Whitney) provides a two-sample comparison of the pseudomedians in the two `treat` groups in terms of `temp_drop`. This is called a **rank sum** test, rather than the Wilcoxon **signed rank** test that is used for inference about a single sample. Here's the resulting 90% confidence interval for the difference in pseudomedians.

```{r}
wt <- sepsis %$% wilcox.test(temp_drop ~ treat,
                       conf.int = TRUE, conf.level = 0.90,
                       alt = "two.sided") 

wt

tidy(wt)
```

